fsOGTT Pre- and Post-ETI

### 2025-11-07 Christine Chan/CF Registry/cf_registry_ROC_analysis.qmd
updated plot for Christine
updates for Christine

```{r setup}
#| include: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(gtsummary)
library(gt)
library(pROC)
library(ggrepel)
home_dir <- switch(
    Sys.info()["sysname"],
    "Darwin" = "/Users/tim/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF/Christine Chan/CF Registry",
    "Windows" = "C:/Users/tim/OneDrive - The University of Colorado Denver/Vigers/CF/Christine Chan/CF Registry",
    "Linux" = "/home/tim/OneDrive/Vigers/CF/Christine Chan/CF Registry"
)
github_dir <- switch(
    Sys.info()["sysname"],
    "Darwin" = "/Users/timvigers/GitHub",
    "Windows" = "C:/Users/Tim/GitHub",
    "Linux" = "/home/tim/GitHub"
)
knitr::opts_knit$set(root.dir = home_dir)
set.seed(1017)
```

@@

# Get BMI group_by
annual$bmi_cat <- apply(annual, 1, function(r) {
    bmi <- as.numeric(r["A_bmivalue"])
    bmi_perc <- as.numeric(r["A_bmipercentile"])
annual$bmi_cat <- relevel(annual$bmi_cat, ref = "Target/Goal")

@@

            bmi_cat,

@@

        bmivalue,
        bmipercentile,
        bmi_cat,

@@

>>>>

```{r setup}
#| include: false
library(dplyr)
library(tidyr)
library(ggplot2)
library(gtsummary)
library(gt)
library(pROC)
library(ggrepel)
library(childsds)
library(labelled)
home_dir <- switch(
  Sys.info()["sysname"],
  "Darwin" = "/Users/tim/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF/Christine Chan/CF Registry",
  "Windows" = "C:/Users/tim/OneDrive - The University of Colorado Denver/Vigers/CF/Christine Chan/CF Registry",
  "Linux" = "/home/tim/OneDrive/Vigers/CF/Christine Chan/CF Registry"
)
github_dir <- switch(
  Sys.info()["sysname"],
  "Darwin" = "/Users/timvigers/GitHub",
  "Windows" = "C:/Users/Tim/GitHub",
  "Linux" = "/home/tim/GitHub"
)
knitr::opts_knit$set(root.dir = home_dir)
set.seed(1017)
```

@@

@@

@@

@@

# Calculate BMI percentiles
roc_df$bmi = roc_df$weight / ((roc_df$height / 100)^2)
roc_df$bmi_perc = sds(
  value = roc_df$bmi,
  age = roc_df$Age_YrEnd,
  sex = roc_df$Gender,
  item = "bmi",
  type = "perc",
  male = "M",
  female = "F",
  ref = cdc.ref
) *
  100
# Raw values for adults
roc_df$bmi_adult = roc_df$bmi
roc_df$bmi_adult[roc_df$Age_YrEnd > 20] = NA
# Get BMI group
roc_df$bmi_cat <- apply(roc_df, 1, function(r) {
  bmi <- as.numeric(r["bmi"])
  bmi_perc <- as.numeric(r["bmi_perc"])
roc_df$bmi_cat <- relevel(roc_df$bmi_cat, ref = "Target/Goal")
# Labels
var_label(roc_df$bmi) = "BMI (all participants)"
var_label(roc_df$bmi_perc) = "BMI percentile (age <= 20)"
var_label(roc_df$bmi_adult) = "BMI value (age > 20)"
var_label(roc_df$bmi_cat) = "BMI category"

### 2025-11-07 Christine Chan/EnVision CF/fsOGTT_pre_post_ETI.qmd
updates for Christine

```{r setup}
#| include: false
library(tidyverse)
library(readxl)
library(lmerTest)
library(performance)
library(gtsummary)
library(emmeans)
home_dir <- switch(Sys.info()["sysname"],
    "Darwin" = "/Users/tim/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF/Christine Chan/EnVision CF",
    "Linux" = "/home/timvigers/OneDrive/Vigers/CF/Christine Chan/EnVision CF"
)
github_dir <- switch(Sys.info()["sysname"],
    "Darwin" = "/Users/timvigers/Documents/GitHub",
    "Linux" = "/home/timvigers/Documents/GitHub"
)
knitr::opts_knit$set(root.dir = home_dir)
```

```{r data cleaning}
#| message: false
# Import cleaned EnVision data
df <- read.csv("./Data_Clean/envision_analysis_dataset.csv", na.strings = "")
# Import GEM data
gem <- read.csv(
    "./Data_Raw/CGMInHealthyControls-FsOGTTDataForMonique_DATA_2025-04-10_0903.csv",
    na.strings = ""
)
# ID linking from Monique
ids <- read.csv(
    "./Data_Clean/TV_ETI logitudinal patient list.csv",
    na.strings = ""
)
ids <- ids %>%
    select(EnVision.ID, GEM.ID) %>%
    rename(`EnVision ID` = EnVision.ID, `GEM ID` = GEM.ID)
ids <- ids[!is.na(ids$`GEM ID`), ]
# Get pre-/post-ETI
df$eti_start <- ifelse(df$corrector___4 == 1, df$corrector_start_date, NA)
df <- df %>%
    group_by(study_id) %>%
    fill(eti_start, .direction = "downup")
# Manually correct a few per Monique
df$eti_start[df$study_id == "CC0001"] <- "2021-11-24"
df$eti_start[df$study_id == "CC0010"] <- "2023-01-16"
df$eti_start[df$study_id == "CC0022"] <- "2019-12-06"
df$eti_start[df$study_id == "CC0023"] <- "2020-01-09"
df$eti_start[df$study_id == "CC0026"] <- "2020-12-04"
df$eti_start[df$study_id == "CC0059"] <- "2021-06-23"
df$pre_post <-
    factor(df$Date >= df$eti_start, levels = c(F, T), labels = c("Pre", "Post"))
df$pre_post[df$study_id == "CC0012"] <- "Pre"
df$pre_post[df$study_id == "CC0017"] <- "Pre"
df$pre_post[df$study_id == "CC0028"] <- "Pre"
# Select columns for EnVision per Monique and combine race and ethnicity
df <- df %>%
    mutate(gem_id = NA, study = "EnVision CF") %>%
    select(
        study_id,
        gem_id,
        study,
        Date,
        age_visit,
        sex,
        origin_race,
        ethnicity,
        CFTR,
        pancreatic_status,
        corrector___4,
        pre_post,
        bmi,
        fev1,
        Diagnosis,
        Glucose_0:Glucose_180,
        Insulin_0:Insulin_180,
        C.Peptide_0:C.Peptide_180
    ) %>%
    unite(race_ethnicity, c(origin_race, ethnicity)) %>%
    rename(ETI = corrector___4)
df$race_ethnicity <- factor(
    df$race_ethnicity,
    levels = c(
        "American Indian or Alaska Native (N. S. & Central America)_Hispanic/Latino",
        "American Indian or Alaska Native (N. S. & Central America)_Not Hispanic/Latino",
        "Black or African American (Africa)_N/A",
        "Black or African American (Africa)_Not Hispanic/Latino",
        "Declined_Hispanic/Latino",
        "Declined_Not Hispanic/Latino",
        "NA_NA",
        "Native Hawaiian or Other Pacific Islander (Hawaii, Guam, Samoa, Pacific Islands)_Not Hispanic/Latino",
        "White (Europe, Middle East, North Africa)_Hispanic/Latino",
        "White (Europe, Middle East, North Africa)_N/A",
        "White (Europe, Middle East, North Africa)_Not Hispanic/Latino"
    ),
    labels = c(
        "Hispanic",
        "Native American",
        "Black/African American",
        "Black/African American",
        "Hispanic",
        "Other/Multiple",
        "Other/Multiple",
        "Other/Multiple",
        "Hispanic",
        "White",
        "White"
    )
)
df$pancreatic_status <- factor(
    df$pancreatic_status,
    levels = 1:2,
    labels = c("Sufficient", "Insufficient")
)
df$ETI <- factor(df$ETI, levels = 0:1, labels = c("No", "Yes"))
# Convert C-peptide level from pmol/L to ng/mL (multiply by 0.003)
df[, grep("C.Peptide_.*", colnames(df))] <-
    lapply(df[, grep("C.Peptide_.*", colnames(df))], function(c) {
        c * 0.003
    })
# Select the same columns from GEM
gem <- gem %>%
    mutate(ETI = 0, pre_post = "Pre", study_id = NA, study = "GEM-CF") %>%
    select(
        study_id,
        subject_id,
        study,
        date_of_visit,
        age_at_visit_1,
        gender,
        race_ethnicity,
        cf_genotype,
        cf_pancreatic,
        ETI,
        pre_post,
        average_bmi,
        fev1_percent,
        group_category,
        lab_ogtt_fasting,
        min_10_glucose,
        min_30_glucose,
        lab_ogtt_1_hour_glucose,
        min_90_glucose,
        lab_ogtt_2_hour_glucose,
        min_150_glucose,
        min_180_glucose,
        min_0_insulin,
        min_10_insulin,
        min_30_insulin,
        min_60_insulin,
        min_90_insulin,
        min_120_insulin,
        min_150_insulin,
        min_180_insulin,
        min_0_c_peptide,
        min_10_c_peptide,
        min_30_c_peptide,
        min_60_c_peptide,
        min_90_c_peptide,
        min_120_c_peptide,
        min_150_c_peptide,
        min_180_c_peptide
    ) %>%
    rename(
        gem_id = subject_id,
        Date = date_of_visit,
        age_visit = age_at_visit_1,
        sex = gender,
        CFTR = cf_genotype,
        pancreatic_status = cf_pancreatic,
        bmi = average_bmi,
        fev1 = fev1_percent,
        Diagnosis = group_category,
        min_0_glucose = lab_ogtt_fasting,
        min_60_glucose = lab_ogtt_1_hour_glucose,
        min_120_glucose = lab_ogtt_2_hour_glucose
    ) %>%
    rename_with(
        ~ paste0("Glucose_", parse_number(.)),
        min_0_glucose:min_180_glucose
    ) %>%
    rename_with(
        ~ paste0("Insulin_", parse_number(.)),
        min_0_insulin:min_180_insulin
    ) %>%
    rename_with(
        ~ paste0("C.Peptide_", parse_number(.)),
        min_0_c_peptide:min_180_c_peptide
    )
# GEM factors
gem$sex <- factor(gem$sex, levels = 1:2, labels = c("Male", "Female"))
gem$race_ethnicity <- factor(
    gem$race_ethnicity,
    levels = 1:6,
    labels = c(
        "White",
        "Black/African American",
        "Asian",
        "Native American",
        "Hispanic",
        "Other/Multiple"
    )
)
gem$CFTR <- factor(
    gem$CFTR,
    levels = c(
        "2184delA/M1101K",
        "2585delT/G551D",
        "3849+10kb C>T/F508del",
        "508del/508del",
        "F508del/1078delA",
        "f508DEL/1154INStc",
        "F508del/1154insTC",
        "F508del/1811 + 1634A > G",
        "F508del/1811+1634>G",
        "F508del/2183delAA->G",
        "F508del/2183delAA>G",
        "F508del/2184insA",
        "F508del/2711delT",
        "F508del/2789+5G>A",
        "F508del/27895GA",
        "F508del/3659delC",
        "F508del/3849+10kbC>T",
        "F508del/394delTT",
        "F508del/5492F",
        "F508del/A561E",
        "F508del/E60X",
        "f508del/F508del",
        "F508del/F508del",
        "F508del/G542X",
        "F508del/G551D",
        "F508del/N1303K",
        "F508del/Q98R",
        "F508del/R1162X",
        "F508del/R117H 9T",
        "F508del/R117H 9T/5T",
        "F508del/R553X",
        "F508del/R709X",
        "F508del/S492F",
        "F508del/W1282X",
        "F5o8del/F508del",
        "G542X/1078delT",
        "G542X/W1089X",
        "G551D/W846X",
        "I507del/G551D",
        "N1303K/1154insTC",
        "N1303K/2789 + 5G > A",
        "N1303K/2789+5G>A",
        "R117H 5t/G576A and R668C on exon 13",
        "R117H5t/G576A and R668C on exon 13",
        "V520F/3791delC",
        "W1089X/S492F",
        "W1090X/S492F"
    ),
    labels = c(
        "Other/Other",
        "Other/Other",
        "Other/Other",
        "F508del homozygous",
        "F508del heterozygous",
        "F508del heterozygous",
        "F508del heterozygous",
        "F508del heterozygous",
        "F508del heterozygous",
        "F508del heterozygous",
        "F508del heterozygous",
        "F508del heterozygous",
        "F508del heterozygous",
        "F508del heterozygous",
        "F508del heterozygous",
        "F508del heterozygous",
        "F508del heterozygous",
        "F508del heterozygous",
        "F508del heterozygous",
        "F508del heterozygous",
        "F508del heterozygous",
        "F508del homozygous",
        "F508del homozygous",
        "F508del heterozygous",
        "F508del heterozygous",
        "F508del heterozygous",
        "F508del heterozygous",
        "F508del heterozygous",
        "F508del heterozygous",
        "F508del heterozygous",
        "F508del heterozygous",
        "F508del heterozygous",
        "F508del heterozygous",
        "F508del heterozygous",
        "F508del homozygous",
        "Other/Other",
        "Other/Other",
        "Other/Other",
        "Other/Other",
        "Other/Other",
        "Other/Other",
        "Other/Other",
        "Other/Other",
        "Other/Other",
        "Other/Other",
        "Other/Other",
        "Other/Other"
    )
)
gem$pancreatic_status <- factor(
    gem$pancreatic_status,
    levels = 1:2,
    labels = c("Sufficient", "Insufficient")
)
gem$ETI <- factor(gem$ETI, levels = 0:1, labels = c("No", "Yes"))
gem$Diagnosis <- factor(
    gem$Diagnosis,
    levels = 1:4,
    labels = c("Healthy Control", "NGT", "AGT", "CFRD")
)
df$Diagnosis <- factor(
    df$Diagnosis,
    levels = c("NGT", "IFG", "IGT", "INDET", "CFRD"),
    labels = c("NGT", "AGT", "AGT", "AGT", "CFRD")
)
# IDs
gem$study_id <- ids$`EnVision ID`[match(gem$gem_id, ids$`GEM ID`)]
df$gem_id <- ids$`GEM ID`[match(df$study_id, ids$`EnVision ID`)]
# Limit GEM per Monique's list
gem <- gem %>% filter(gem_id %in% ids$`GEM ID`)
# Merge, keep only pre/post, sort by ID and date
df <- rbind(df, gem) %>% filter(!is.na(pre_post))
df$pre_post <- factor(df$pre_post, levels = c("Pre", "Post"))
df <- df %>% arrange(study_id, Date)
# Calculate ODI, etc.
# Write for checking
write.csv(
    df,
    file = "./Data_Clean/fsOGTT_pre_post_ETI.csv",
    row.names = F,
    na = ""
)
```

>>>>

```{r setup}
#| include: false
library(tidyverse)
library(gtsummary)
library(labelled)
library(plotly)
library(lmerTest)
home_dir <- switch(
  Sys.info()["sysname"],
  "Darwin" = "/Users/tim/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF/Christine Chan/EnVision CF",
  "Windows" = "C:/Users/Tim/OneDrive - The University of Colorado Denver/Vigers/CF/Christine Chan/EnVision CF",
  "Linux" = "/home/tim/OneDrive/Vigers/CF/Christine Chan/EnVision CF"
)
github_dir <- switch(
  Sys.info()["sysname"],
  "Darwin" = "/Users/tim/Documents/GitHub",
  "Windows" = "C:/Users/Tim/Documents/GitHub",
  "Linux" = "/home/tim/Documents/GitHub"
)
knitr::opts_knit$set(root.dir = home_dir)
```

```{r data cleaning}
# Import cleaned data
load("./Data_Clean/fsOGTT_pre_post_ETI.RData")
# Subset to people with more than one pre-ETI fsOGTT
multi_pre_df = df |>
  group_by(study_id) |>
  mutate(
    n_pre = sum(pre_post == "Pre"),
    time_diff_first_last = as.numeric(difftime(
      last(Date),
      first(Date),
      units = "days"
    ))
  ) |>
  filter(n_pre > 1)
var_label(multi_pre_df$n_pre) = "Number of fsOGTT pre-ETI"
var_label(multi_pre_df$time_diff_first_last) = "Days From First to Last fsOGTT"
# Save for Christine and Monique
write.csv(
  multi_pre_df,
  file = "./Data_Clean/fsOGTT_pre_post_ETI.csv",
  row.names = F,
  na = ""
)
```

# Participants with > 1 pre-ETI fsOGTT

## Table 1: Participant characteristics at first fsOGTT

```{r}
#| label: tbl-1
multi_pre_df |>
  group_by(study_id) |>
  filter(row_number() == 1) |>
  ungroup() |>
  select(
    n_pre,
    time_diff_first_last,
    sex,
    age_visit,
    bmi,
    bmi_adult,
    bmi_perc,
    fev1,
    CFTR,
    pancreatic_status,
    Diagnosis,
    Insulin_0,
    Glucose_0,
    igi,
    iAUC30ins_over_gluc,
    iAUC120ins_over_gluc,
    iAUC180ins_over_gluc,
    odi_igi,
    odi_iauc
  ) |>
  tbl_summary(type = c("Insulin_0") ~ "continuous")
```

## IGI (delta 30min insulin/delta 30min glucose)

```{r}
#| warning: false
# Plot
p = ggplot(
  multi_pre_df,
  aes(x = years_from_eti_start, y = igi, group = study_id, color = study_id)
) +
  geom_point() +
  geom_line() +
  theme_bw()
ggplotly(p)
# Model
m = lmer(
  igi ~ years_from_eti_start_model + slope_change + (1 | study_id),
  data = multi_pre_df
)
m |>
  tbl_regression(
    exponentiate = F,
    estimate_fun = label_style_sigfig(digits = 4),
    pvalue_fun = label_style_pvalue(digits = 2),
  )
```

## iAUC30ins/iAUC30gluc

```{r}
#| warning: false
# Plot
p = ggplot(
  multi_pre_df,
  aes(
    x = years_from_eti_start,
    y = iAUC30ins_over_gluc,
    group = study_id,
    color = study_id
  )
) +
  geom_point() +
  geom_line() +
  theme_bw()
ggplotly(p)
# Model
m = lmer(
  iAUC30ins_over_gluc ~ years_from_eti_start_model +
    slope_change +
    (1 | study_id),
  data = multi_pre_df
)
m |>
  tbl_regression(
    exponentiate = F,
    estimate_fun = label_style_sigfig(digits = 4),
    pvalue_fun = label_style_pvalue(digits = 2),
  )
```

## iAUC120ins/iAUC120gluc

```{r}
#| warning: false
# Plot
p = ggplot(
  multi_pre_df,
  aes(
    x = years_from_eti_start,
    y = iAUC120ins_over_gluc,
    group = study_id,
    color = study_id
  )
) +
  geom_point() +
  geom_line() +
  theme_bw()
ggplotly(p)
# Model
m = lmer(
  iAUC120ins_over_gluc ~ years_from_eti_start_model +
    slope_change +
    (1 | study_id),
  data = multi_pre_df
)
m |>
  tbl_regression(
    exponentiate = F,
    estimate_fun = label_style_sigfig(digits = 4),
    pvalue_fun = label_style_pvalue(digits = 2),
  )
```

## iAUC180ins/iAUC180gluc

```{r}
#| warning: false
# Plot
p = ggplot(
  multi_pre_df,
  aes(
    x = years_from_eti_start,
    y = iAUC180ins_over_gluc,
    group = study_id,
    color = study_id
  )
) +
  geom_point() +
  geom_line() +
  theme_bw()
ggplotly(p)
# Model
m = lmer(
  iAUC180ins_over_gluc ~ years_from_eti_start_model +
    slope_change +
    (1 | study_id),
  data = multi_pre_df
)
m |>
  tbl_regression(
    exponentiate = F,
    estimate_fun = label_style_sigfig(digits = 4),
    pvalue_fun = label_style_pvalue(digits = 2),
  )
```

## ODI (IGI*Matsuda)

```{r}
#| warning: false
# Plot
p = ggplot(
  multi_pre_df,
  aes(x = years_from_eti_start, y = odi_igi, group = study_id, color = study_id)
) +
  geom_point() +
  geom_line() +
  theme_bw()
ggplotly(p)
# Model
m = lmer(
  odi_igi ~ years_from_eti_start_model + slope_change + (1 | study_id),
  data = multi_pre_df
)
m |>
  tbl_regression(
    exponentiate = F,
    estimate_fun = label_style_sigfig(digits = 4),
    pvalue_fun = label_style_pvalue(digits = 2),
  )
```

## ODI [(iAUC30ins/iAUC30gluc)*Matsuda]

```{r}
#| warning: false
# Plot
p = ggplot(
  multi_pre_df,
  aes(
    x = years_from_eti_start,
    y = odi_iauc,
    group = study_id,
    color = study_id
  )
) +
  geom_point() +
  geom_line() +
  theme_bw()
ggplotly(p)
# Model
m = lmer(
  odi_iauc ~ years_from_eti_start_model + slope_change + (1 | study_id),
  data = multi_pre_df
)
m |>
  tbl_regression(
    exponentiate = F,
    estimate_fun = label_style_sigfig(digits = 4),
    pvalue_fun = label_style_pvalue(digits = 2),
  )
```
